Symmetric Peer-to-Peer Applications

### 2024-10-25 rita-24/response.md

We accommodate all clarifications and answers in a final version of the paper.
To help the reviewers find the modifications of interest, we provide a link
with a *diff* between the versions, as follows:

```
https://
```

Based on the overall comments, our focus on the final version was to improve
the section on related work, and to make the limitations of our proposal more
explicit.

For that matter, we expanded Section "2. Related Work" by splitting it in two
subsections
    "2.1. Symmetric Distributed Applications" and
    "2.2. Software Time Machines".
We also rewrote Section "3.2.4. Middleware Summary" to discuss its limitations.

Next, we address each individual comment from all reviewers as follows...
(Some comments might be reordered.)

@@

- 1. Introduction:

> "We target small peer-to-peer networks, in which nodes are only a few hops
> away from each other. This ensures that events can span the whole network in
> a reasonable time to preserve the real-time behavior of applications. We also
> assume that peers are non-malicious in the sense that they do not generate
> erroneous events."

- 3.2.1. Event Dissemination:

> "Considering the peer-to-peer network as a whole, each event packet is
> replicated to all neighbours of all peers. This results in a theoretical
> limit of quadratic messages if all peers were connected to all peers. In
> contrast, as discussed in Section 2.1, a centralized solution only requires a
> packet to be sent to a server, which then broadcasts it to all clients."

@@

- 3.2.4. Middleware Summary:

We discuss the middleware limitaions as follows:
contiguous unique identifiers, deterministic API, non-malicious peers, and
static memory footprint.
The text is too long to fit here and can be found in the paper or link above.

@@

> "Regarding the static limitation, we mentioned the possibility of a custom
> allocator, which is illustrated as follows:"

```
struct {
    // FIXED REGION
    int x,  y;   // position
    int dx, dy;  // direction speed
    // VARIABLE REGION
    char heap[HEAP_SIZE];
} G;

T* obj = my_malloc(G.heap, <size>);
<...>
my_free(obj);
```

> "The variable region is still technically static, but is dynamically reshaped
> internally with the custom allocator `my_alloc` & `my_free`."

@@

> "The step delay is dynamically adjusted such that the whole time travel loop
> takes at most 1 second to complete.
> Note that during this period, the application is frozen, and new inputs are
> enqueued for further processing."

### Comment A.6

> <...>
> Suggestion: if possible please consider providing a more formal description
> of the time synchronization protocol, including any guarantees or assumptions
> about clock drift and network delays.

Even though we consider that this suggestion is relevant, we believe that a
formal description of the protocol with its implications requires a dedicated
work on its own.

@@

Regarding the experimental evaluation (comments A.7-A.9):

### Comment A.7

> <...>
> The experimental evaluation could be more extensive, including more
> diverse scenarios *(see A.8)* and a comparison with existing approaches

At the end of Section "4. Evaluation", we now compare the existing approaches
discussed in Section "2.1. Symmetric Distributed Applications" with respect to
the three aspects that we consider in our evaluation (i.e., recurrence of time
travels, real-time pace of peers, correctness of unstable peers).
The text is too long to fit here and can be found in the paper or link above.

@@

### Comment A.8

> <...>
> The evaluation uses a fixed network topology. How would the middleware's
> performance be affected in scenarios with highly dynamic topologies, where
> peers frequently join, leave, or experience intermittent connectivity?
> Have you considered any mechanisms to adapt to such changes?

Even though the network topology is fixed, it includes three typical variations
of (a) straight bus lines, (b) rings, and (c) connected topologies.
They all coexist in the same deployment with peers leaving and joining the
network.
In item "(c) Correctness of unstable peers", we make all 21 nodes to switch
between online and offline for the same amount of time, including the peers in
the middle of the network.
This churn rate of 50% periodically creates full partitions in the peer-to-peer
network, making the two opposites incommunicable for a considerable amount of
time.

@@

- 2.1 Symmetric Distributed Applications: discussion on the selected works,
    and comparison with our proposal.
- 2.2 Software Time Machines: expanded and improved discussion on time
    machines.

### Comment J.2

> ii) The introduction on the related works makes it appear like there are
> ideally 3 subsections: (a) how the network is organized, (b) how global time
> is determined, and (c) how events are propagated and applied.
> However, this three aspects are actually related to the discussion on the
> selected works compared to the proposed solution.

Now, to make clear that we have only two subsections, we briefly introduce 2.1
and 2.2 with a paragraph at the beginning of Section 2 (before 2.1 and 2.2):

@@

Done, as detailed in J.1 and J.2.

### Comment J.4

> iii) The discussion does not have a very smooth transition. The punchline of
> the related works section should ideally be the proposal of this work, which
> is brought out more clearly when discussed at the end of the related works to
> show how all the gaps identified will be solved.

To guide our discussion, we now explain how the text is organized at the
beginning of Section 2.1:

@@

> <...>
> "Figure 1 compares three selected works regarding these aspects. We now
> discuss these works with respect to the three aspects, while comparing with
> our approach. At the end of this section, we also provide initial details on
> how our middleware works."

### Comment J.5

> iv) The discussion on related work on time machines is rather abrupt.
> It does not have a proper introduction to the problem of time machines in
> relation to this work (or in general for that matter) but directly proceeds
> into an analysis of Fusion and Braid.
> --> What is aimed by this section? Answering this can help find a way of
> introducing the time machines aspect well to the reader.

The discussion was indeed abrupt and out of context, so we reshaped the new
subsection 2.2. We included two new paragraphs at the very beginning, in which
we discuss
    (a) what a software time machine is,
    (b) in which scenarios it appears,
    (c) what are the challenges to overcome, and
    (d) what are the common implementation techniques.
The text is too long to fit here and can be found in the paper or link above.

@@

Regarding the protocol, middleware, and peer-to-peer concepts (comments
J.7-J.11):

### Comment J.7

> i) As a proof-of-concept, it will be important to have p2p
> model/framework to give a clear picture to an expert reader and explain to a
> lay reader the place of this middleware in the overall p2p application
> model/framework. Has this been considered?

We now include a paragraph at the beginning of Section "3. The Middleware &
Programming API" describing the P2P framework or, in other words, what we
exactly mean by P2P:

> "The peers form a dynamic unstructured peer-to-peer network [3], and
> communicate only indirectly to each other. Events are flooded in the network
> graph via gossiping, i.e.: when a peer generates an event, it communicates to
> its neighbours, which communicate to their neighbours, and so on. Note that
> all peers execute the exactly same application and middleware, with no
> differences with respect to their roles and physical resources. Therefore, we
> can describe our peer-to-peer network as follows: (a) all peers have the same
> role and run the same software; (b) any peer can join or leave at any time;
> (c) events are only communicated with direct neighbours."

@@

> "Peers in the application form a dynamic network graph and communicate only
> with direct neighbours, as in typical unstructured peer-to-peer networks.
> Events are flooded in the graph and are triggered locally with a small delay
> to compensate the network latency."

What we meant is that the same effect of a broadcast is reached after the
events are flooded in the network, but only through direct neighbour
communication.
Nevertheless, we no longer use this loosened notion of a broadcast.

We now also emphasize this distinction from centralized solutions (which do
rely on broadcasts) in Section "2. Related Work":

> "Note how the central server is the only participant with knowledge about
> clients, which never communicate directly among themselves.
> This implies that the server must never fail, and also that periodic full
> broadcasts are the only possibility of spanning the whole network.
> In this work, we argue that in a peer-to-peer alternative any node can fail,
> and that broadcasts can be replaced by optimistic flooding."
> <...>
> "Note, that it is not necessary that all peers communicate directly with all
> other peers, but only indirectly via unstructured flooding."
> <...>

### 2024-12-06 rita-24/response.md

# Response Letter

We would like to thank the reviewers for the thoughtful comments, questions,
and criticism.

We accommodate all clarifications and answers in a final version of the paper.
To help the reviewers find the modifications of interest, we provide a link
with a *diff* between the versions, as follows:

```
https://github.com/fsantanna-no/p2p-tml-paper/compare/690245...main
```

@@

- 3.2.1. Event Dissemination:

> "Considering the peer-to-peer network as a whole, each event packet is
> replicated to all neighbours of all peers. This results in a theoretical
> limit of quadratic messages if all peers were connected to all peers. In
> contrast, as discussed in Section 2.1, a centralized solution only requires a
> packet to be sent to a server, which then broadcasts it to all clients."

- 3.2.4. Middleware Summary:

We discuss the middleware limitations as follows:
contiguous unique identifiers, deterministic API, non-malicious peers, and
static memory footprint.
The text is too long to fit here and can be found in the paper or link above.

@@

At the end of Section "4. Evaluation", we now compare the existing approaches
discussed in Section "2.1. Symmetric Distributed Applications" with respect to
the three aspects that we consider in our evaluation (i.e., recurrence of time
travels, real-time pace of peers, correctness of unstable peers).
The text is too long to fit here and can be found in the paper or link above.

### Comment A.8

> <...>
> The evaluation uses a fixed network topology. How would the middleware's
> performance be affected in scenarios with highly dynamic topologies, where
> peers frequently join, leave, or experience intermittent connectivity?
> Have you considered any mechanisms to adapt to such changes?

Even though the network topology is fixed, it includes three typical variations
of (a) straight bus lines, (b) rings, and (c) connected topologies.
They all coexist in the same deployment with peers leaving and joining the
network.
In item "(c) Correctness of unstable peers", we make all 21 nodes to switch
between online and offline for the same amount of time, including the peers in
the middle of the network.
This churn rate of 50% periodically creates full partitions in the peer-to-peer
network, making the two opposites incommunicable for a considerable amount of
time.
Therefore, our unstructured design does provide the necessary mechanisms to
adapt to such changes, since peers joins and leaves only impact direct
neighbours.
Our evaluation takes high churn rates into consideration.

@@

### Comment J.4

> iii) The discussion does not have a very smooth transition. The punchline of
> the related works section should ideally be the proposal of this work, which
> is brought out more clearly when discussed at the end of the related works to
> show how all the gaps identified will be solved.

To guide our discussion, we now explain how the text is organized at the
beginning of Section 2.1:

> <...>
> "Figure 1 compares three selected works regarding these aspects, which we
> discuss and contrast with our approach next. At the end of this section, we
> also provide initial details on how our middleware works.

@@

### Comment J.7

> i) As a proof-of-concept, it will be important to have p2p
> model/framework to give a clear picture to an expert reader and explain to a
> lay reader the place of this middleware in the overall p2p application
> model/framework. Has this been considered?

We now include a paragraph at the beginning of Section "3. The Middleware &
Programming API" describing the P2P framework or, in other words, what we
exactly mean by P2P:

> "The peers form a dynamic unstructured peer-to-peer network [3], and
> communicate mostly indirectly to each other. Events are flooded in the
> network graph via gossiping, i.e.: when a peer generates an event, it
> communicates to its neighbours, which communicate to their neighbours, and so
> on. Note that all peers execute the exactly same application and middleware,
> with no differences with respect to their roles and physical resources.
> Therefore, we can describe our peer-to-peer network as follows: (a) all peers
> have the same role and run the same software; (b) any peer can join or leave
> at any time; (c) events are only communicated with direct neighbours."

@@

> "As an optimization, the middleware takes periodic snapshots locally *at all
> peers* to avoid full simulation."
> <...>
> "In addition, a peer that rejoins the network recovers from a past snapshot
> it already holds, thus avoiding full simulation."

### Comment J.10

> iv) The simulations were done on the same machine. This means that there is
> no actual considerations due to the effect of network challenges. Is this the
> case?

The evaluation requires a synchronized global clock that can measure
differences in the order of a few milliseconds.
This can be done reliably in a single machine, but not trivially across the
network with the precision that we needed.
We use Linux's NetEm to emulate network latency, but do not consider other
network challenges.
That being said, latency is the only network dependent parameter that we
currently evaluate.
